Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-07-31 latex/old.tex
Intro ready

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  nyberg2012, fedorov1997, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy problem: inferring
the initial condition of the 1D heat equation (details in the
supplementary material). In Fig.~\ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. We refer to this intriguing
phenomenon as \emph{measurement clusterization} \cite{Ucinski05}, and
we refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

%% first: blue, second: red, third: green, fourth: black,
%% fifth:magenta
\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{figs/example.png}
    \caption{Measurement clusterization for optimal designs when
      inverting for the initial condition of the 1D heat equation (see
      supplementary material for details). Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain \(\Omega = [0, 1]\)
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid with the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

@@

\item \label{q:mitigate} \textbf{Why does imposing correlations
  between observations alleviate measurement clusterization?} In
  Section \ref{section:non vanishing}, we rigorously demonstrate the
  role of model error in mitigating clusterization, thereby
  corroborating earlier observations made by various researchers.

@@

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let \(\hilp\) and \(\hilo\) separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively), and let
\(\fwd: \hilp \to \hilo\) the \emph{forward operator}. The forward
operator \(\fwd\) is assumed linear and strongly smoothing (the heat
operator of the 1D heat equation introduced in Section
\ref{section:intro} is a prime example). Take a Gaussian prior
\(\param \sim \pr = \normal(\prmean ,\prcov)\) with some appropriate
covariance operator \(\prcov\) on \(\hilp\) \cite{Stuart10}. Note that
\(\fwd \prcov \fwd^*\) is the prior covariance in \(\hilo\)
\cite{Stuart10}, and as such, \(\fwd \prcov \fwd^*\) is assumed
invertible --- an assumption which will be used below (if \(\fwd\) has
a nontrivial kernel we utilize Occam's Razor and ignore said kernel
altogether). Measurements are taken via a \emph{measurement operator}
\(\obs\). It is common for the measurement and forward operators to be
merged \(\tmp := \obs \fwd\) \cite{AlexanderianGloorGhattas14}, but
the analysis carried out in the following sections requires that
\(\fwd\) and \(\obs\) are explicitly separated as in
\cite{attia2022stochastic, cvetkovic2023choosing}.  \(\obs \in (
\hilo^* )^m\), where \(m\) is the number of measurements
taken. Entries \(\meas_j, j=1,\dots,m\) of the observation operator
\(\obs\) are called \emph{measurements}:

@@

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error \(\eps' \sim
\normal(0,\modcov)\) with \(\modcov\) a covariance operator. Observation
error is denoted \(\eps \sim \normal(0, \sigma^2 I_m)\), with \(I_m \in
\mathbb{R}^{m \times m}\) the identity. Both error terms and the prior
are assumed independent of each other. Thus, data is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

@@

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
Taking \(\modcov = 0\) is a common practice
\cite{tarantola2005,Kaipio2005,Vogel02} and then \(\Sigma =
\sigma^2I_m\) is a scalar matrix which does not depend on \(\obs\).

@@

\subsection{Bayesian D-Optimal Designs in Infinite Dimensions}\label{subsec:D optimal design} 
A Bayesian D-optimal design maximizes expected Kullback-Leibler (KL)
divergence between posterior \(\post\) and prior measures \(\pr\). It is
useful to first recall the definition of KL divergence \cite{CoverThomas91}:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}

@@

\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let \(\pr = \normal(\prmean,\prcov)\) be a Gaussian prior on \(\hilp\)
  and let \(\post = \normal(\postmean,\postcov)\) the posterior measure
  on \(\hilp\) for the Bayesian linear inverse problem \(\data = \obs
  \fwd\param + \obs \eps' + \eps\) discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

@@

%% It is important to note that since \(\obs\) is finite-rank,
%% \(\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}\) is
%% trace-class.
\begin{definition}\label{def:d optimality}
  We say \(\obs^{\star}\) is \emph{D-optimal} if \(\obs^{\star} =
  \argmax_{\obs} \tar(\obs)\), where entries of \(\obs \in (\hilo^*)^m\)
  are constrained to some allowed set of measurements in \(\hilo^*\).
\end{definition}

Intuition for Theorem \ref{thm:d optimality} can be gained by
considering a Bayesian linear model in finite dimensions, with
Gaussian prior and Gaussian noise. Then, a D-optimal design minimizes
the determinant of the posterior covariance matrix, and this turns out
to be a regularized version of the frequentist D-optimality criterion
\cite{Chaloner1995}. Theorem \ref{thm:d optimality} and Definition
\ref{def:d optimality} carry a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
    %%  &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
We think of \(\prcov\) as constant, so a D-optimal design minimizes a
quantity analogous to the posterior covariance determinant, similarly
to the finite-dimensional case.

@@

%% \subsection{Sequential vs Simultaneous Optimization}\label{subsec:seq vs sim}
%% From defintion \ref{def:d optimality} we wish to characterize solution(s) of the
%% following optimization problem for \(\tar$. %%: (\hilo^*)^m \to \R$:
%% \begin{align}\label{eq:optimization}
%%   \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
%%   = \argmax_{\obs} \frac12 \log \det 
%%   (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
%% \end{align}
%% where \(\obs\) is constrained to some allowed set of observations. We
%% call this problem ``simultaneous optimization'', since all
%% observations are decided on simulatneously.

%% For computational reasons, one may prefer to find the best
%% observations in a sequential manner. Denote
%% \begin{equation}\label{eq:def obs_k}
%%   \obs_k := (\meas_1,\dots, \meas_k)^t,  k\leq m.
%% \end{equation}
%% Sequential optimal design proceeds as follows. Find \(\meas_1\) by
%% maximizing \(\tar(\obs_1)$. Then, keeping \(\meas_1\) fixed --- find
%% \(\meas_2\) as the maximizer of \(\tar(\obs_2)$. Then, find \(\meas_3\) by
%% keeping \(\meas_1,\meas_2\) fixed and taking \(\meas_3\) as the maximizer
%% of \(\tar(\obs_3)$. Continue this way until \(\obs_m = \obs\) is
%% found, where \(m\) is the number of available observations. %% It is
%% %% important to notice that this scheme does not require actually
%% %% observing data --- in \eqref{eq:objective} data is averaged out.

%% The analysis in this paper is conducted for the general simultaneous
%% optimization case. The sequential optimization case is dealt with in
%% section \ref{subsec:clusterization sequential}. It is important to
%% note, however that all conclusions we arrive at for the simultaneous
%% case easily specialize to the sequential case by considering the
%% posterior as the next sequential step's prior.

\section{The Constrained Optimization Problem of D-Optimal Design}\label{section:D and grad}
We seek a formulation of the D-optimal design problem via Lagrange
multipliers. We first find the gradient of $\tar$, then we suggest
unit-norm constraints on $\obs$ and find their gradients. Results of
this section are summarized in Theorem \ref{thm:constrained}. First,
recall that:
\begin{definition}\label{def:var}
  Let $F$ a real valued function of $\obs$. The first variation of $F$
  at $\obs$ in the direction $V$ is:
  \begin{equation*}
    \delta F(\obs) V := \frac{\der}{\der \tau}\Big |_{\tau=0}  F( \obs + \tau V).
  \end{equation*}

@@

\begin{proposition}\label{prop:tar grad}
  The gradient of the D-optimality objective $\tar$ is
  \begin{equation*}
    %% \delta \tar(\obs) V = \tr{V ( I - \modcov \obs^* \Sigma^{-1}\obs )
    %%   \fwd \postcov \fwd^* \obs^* \Sigma^{-1}}.
    \nabla \tar(\obs) = ( I - \modcov \obs^* \Sigma^{-1}\obs ) \fwd
    \postcov \fwd^* \obs^* \Sigma^{-1}
  \end{equation*}
\end{proposition}

@@

  \begin{align}\label{eq:der sig}
    \begin{split}
      \frac{\der}{\der \tau} \Big |_{\tau=0} \Sigma( \obs + \tau V )
      &= \frac{\der}{\der \tau} \Big |_{\tau=0} 
      (\obs + \tau V ) \modcov (\obs + \tau V )^*  + \sigma^2I\\
      % 
      % 
      % 
      &= V \modcov \obs^* + \obs \modcov V^*.
    \end{split}
  \end{align}

  Then, using \eqref{eq:der sig}: 
  \begin{align*}
    0 &= \frac{\der}{\der \tau} \Big |_{\tau=0} I \\
    % 
    % 
    % 
    &= \frac{\der}{\der \tau} \Big |_{\tau=0}
    \left (\Sigma(\obs+\tau V)^{-1} \Sigma(\obs+\tau V) \right ) \\
    % 
    % 
    % 
    &= \frac{\der \Sigma(\obs+\tau V)^{-1}}{\der \tau} \Big |_{\tau=0} \Sigma+
    \Sigma^{-1} \frac{\der \Sigma(\obs+\tau V)}{\der \tau} \Big |_{\tau=0}\\  
    %
    %
    %
    &= \frac{\der \Sigma(\obs+\tau V)^{-1}}{\der \tau} \Big |_{\tau=0} \Sigma+
    \Sigma^{-1} (V\modcov \obs^* + \obs \modcov V^*). 
    %\text{, by \eqref{eq:der sig}. }
  \end{align*}

  Thus:
  \begin{align}\label{eq:der sig inv}
    \frac{\der \Sigma(\obs+\tau V)^{-1}}{\der \tau} \Big |_{\tau=0}  
      &= -\Sigma^{-1} (V \modcov \obs^* + \obs \modcov V^*) \Sigma^{-1}.
    \end{align}
  %

@@

  By Leibniz (product) rule and \eqref{eq:der sig inv}:
  % 
  \begin{align}\label{eq:T}
    \begin{split}
    \delta T(\obs) V 
    &= \frac{\der T(\obs + \tau V)}{\der \tau} \Big |_{\tau=0} \\
    %
    %
    %
    &= V^* \Sigma^{-1} \obs 
    - \obs^*\Sigma^{-1} V\modcov \obs^* \Sigma^{-1}\obs \\
    &\ \ \ - \obs^* \Sigma^{-1} \obs \modcov V^* \Sigma^{-1}\obs
    + \obs^* \Sigma^{-1} V.
    \end{split}
  \end{align}

@@

  \begin{align*}
    %\begin{split}
      \delta \tar(\obs) V 
      :&= \frac{\der}{\der\tau} \Big |_{\tau=0} \tar(\obs + \tau V) \text{ (Definition \ref{def:var})}\\
      % 
      % 
      % 
      &= \frac12 \frac{\der}{\der \tau} \Big |_{\tau=0} \log \det 
      (I + \prcov^{1/2} \fwd^* T(\obs+\tau V)\fwd \prcov^{1/2} ) \text{ (Theorem \ref{thm:d optimality})} \\
      % 
      % 
      % 
      &= \frac12 \tr{( I + \prcov^{1/2} \fwd^* \obs^* \Sigma^{-1}
        \obs\fwd \prcov^{1/2} )^{-1}
        \frac{\der}{\der \tau} \Big |_{\tau=0}
        \prcov^{1/2} \fwd^* T(\obs+\tau V) \fwd \prcov^{1/2}}\ \text{ (Lemma \ref{lemma:lax})} \\
      % 
      % 
      % 
      &= \frac12 \ttr\Big \{ \postcov \fwd^* (V^* \Sigma^{-1} \obs 
      - \obs^*\Sigma^{-1} V\modcov \obs^* \Sigma^{-1}\obs \\
      &\ \ \ - \obs^* \Sigma^{-1} \obs \modcov V^* \Sigma^{-1}\obs 
      + \obs^* \Sigma^{-1} V ) \fwd \Big \}  \text{ (by \eqref{eq:T})} \\
      %
      %
      %
      &= \tr{\postcov \fwd^* ( \obs^* \Sigma^{-1} V -
      \obs^*\Sigma^{-1} V\modcov \obs^* \Sigma^{-1}\obs ) \fwd} \\
      %
      %
      % 
      &= \tr{\postcov \fwd^* \obs^* \Sigma^{-1} V 
      ( I - \modcov \obs^* \Sigma^{-1}\obs ) \fwd} \\
      % 
      %
      %
      &= \tr{V ( I - \modcov \obs^* \Sigma^{-1}\obs )
      \fwd \postcov \fwd^* \obs^* \Sigma^{-1}}.
    %\end{split}
  \end{align*} 
  Recalling Definition \ref{def:var} concludes the proof.
\end{proof}

@@

\begin{proposition}\label{prop:bigger better}
  Let $\obs = (\meas_1,\dots,\meas_m)^t$, $j \in \{1,\dots,m\}$,
  $\sigma^2 > 0$ and $|\zeta| > 1$. Then $\tar(\obs)$ increases if we
  use $\zeta \meas_j$ in $\obs$ instead of $\meas_j$.
\end{proposition}

>>>>

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  nyberg2012, fedorov1997, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy problem: inferring the
initial condition of the 1D heat equation (details in the
supplementary material). In Fig.~\ref{fig:clusterization_illustration}
D-optimal measurement locations are shown for different numbers of
measurements. Notably, for six measurements, a D-optimal design yields
two sets of measurements that are almost indistinguishable from one
another. We refer to this intriguing phenomenon as \emph{measurement
clusterization} \cite{Ucinski05}, and we refer to a design that
exhibits measurement clusterization as a \emph{clustered design}.

%% first: blue, second: red, third: green, fourth: black,
%% fifth:magenta
\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{figs/example.png}
    \caption{Measurement clusterization for optimal designs when
      inverting for the initial condition of the 1D heat equation (see
      supplementary material for details). Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d_optimality}. Measurement locations are
      plotted over the computational domain \(\Omega = [0, 1]\)
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid with the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization_illustration}
\end{figure}

@@

\item \label{q:mitigate} \textbf{Why does imposing correlations
  between observations alleviate measurement clusterization?} In
  Section \ref{section:non_vanishing}, we rigorously demonstrate the
  role of model error in mitigating clusterization, thereby
  corroborating earlier observations made by various researchers.

@@

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract_OED}
Let \(\hilp\) and \(\hilo\) separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively), and let
\(\fwd: \hilp \to \hilo\) the \emph{forward operator}. The forward
operator \(\fwd\) is assumed linear and strongly smoothing (the heat
operator of the 1D heat equation introduced in Section
\ref{section:intro} is a prime example). Take a Gaussian prior
\(\param \sim \pr = \normal(\prmean ,\prcov)\) with some appropriate
covariance operator \(\prcov\) on \(\hilp\) \cite{Stuart10}. Note that
\(\fwd \prcov \fwd^*\) is the prior covariance in \(\hilo\)
\cite{Stuart10}, and as such, \(\fwd \prcov \fwd^*\) is assumed
invertible --- an assumption which will be used below (if \(\fwd\) has
a nontrivial kernel we utilize Occam's Razor and ignore said kernel
altogether). Measurements are taken via a \emph{measurement operator}
\(\obs\). It is common for the measurement and forward operators to be
merged \(\tmp := \obs \fwd\) \cite{AlexanderianGloorGhattas14}, but
the analysis carried out in the following sections requires that
\(\fwd\) and \(\obs\) are explicitly separated as in
\cite{attia2022stochastic, cvetkovic2023choosing}.  \(\obs \in (
\hilo^* )^m\), where \(m\) is the number of measurements
taken. Entries \(\meas_j, j=1,\dots,m\) of the observation operator
\(\obs\) are called \emph{measurements}:

@@

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error \(\eps' \sim
\normal(0,\modcov)\) with \(\modcov\) a covariance operator. Observation
error is denoted \(\eps \sim \normal(0, \sigma^2 I_m)\), with \(I_m \in
\mathbb{R}^{m \times m}\) the identity. Both error terms and the prior
are assumed independent of each other. Thus, data is acquired via
\begin{align}\label{eq:inverse_problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

@@

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov_explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
Taking \(\modcov = 0\) is a common practice
\cite{tarantola2005,Kaipio2005,Vogel02} and then \(\Sigma =
\sigma^2I_m\) is a scalar matrix which does not depend on \(\obs\).

@@

\subsection{Bayesian D-Optimal Designs in Infinite Dimensions}\label{subsec:D_optimal_design} 
A Bayesian D-optimal design maximizes expected Kullback-Leibler (KL)
divergence between posterior \(\post\) and prior measures \(\pr\). It is
useful to first recall the definition of KL divergence \cite{CoverThomas91}:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}

@@

\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d_optimality}
  Let \(\pr = \normal(\prmean,\prcov)\) be a Gaussian prior on \(\hilp\)
  and let \(\post = \normal(\postmean,\postcov)\) the posterior measure
  on \(\hilp\) for the Bayesian linear inverse problem \(\data = \obs
  \fwd\param + \obs \eps' + \eps\) discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

@@

%% It is important to note that since \(\obs\) is finite-rank,
%% \(\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}\) is
%% trace-class.
\begin{definition}\label{def:d_optimality}
  We say \(\obs^{\star}\) is \emph{D-optimal} if \(\obs^{\star} =
  \argmax_{\obs} \tar(\obs)\), where entries of \(\obs \in (\hilo^*)^m\)
  are constrained to some allowed set of measurements in \(\hilo^*\).
\end{definition}

Intuition for Theorem \ref{thm:d_optimality} can be gained by
considering a Bayesian linear model in finite dimensions, with
Gaussian prior and Gaussian noise. Then, a D-optimal design minimizes
the determinant of the posterior covariance matrix, and this turns out
to be a regularized version of the frequentist D-optimality criterion
\cite{Chaloner1995}. Theorem \ref{thm:d_optimality} and Definition
\ref{def:d_optimality} carry a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
    %%  &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
We think of \(\prcov\) as constant, so a D-optimal design minimizes a
quantity analogous to the posterior covariance determinant, similarly
to the finite-dimensional case.

@@

%% \subsection{Sequential vs Simultaneous Optimization}\label{subsec:seq_vs_sim}
%% From defintion \ref{def:d_optimality} we wish to characterize solution(s) of the
%% following optimization problem for \(\tar$. %%: (\hilo^*)^m \to \R$:
%% \begin{align}\label{eq:optimization}
%%   \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
%%   = \argmax_{\obs} \frac12 \log \det 
%%   (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
%% \end{align}
%% where \(\obs\) is constrained to some allowed set of observations. We
%% call this problem ``simultaneous optimization'', since all
%% observations are decided on simulatneously.

%% For computational reasons, one may prefer to find the best
%% observations in a sequential manner. Denote
%% \begin{equation}\label{eq:def_obs_k}
%%   \obs_k := (\meas_1,\dots, \meas_k)^t,  k\leq m.
%% \end{equation}
%% Sequential optimal design proceeds as follows. Find \(\meas_1\) by
%% maximizing \(\tar(\obs_1)$. Then, keeping \(\meas_1\) fixed --- find
%% \(\meas_2\) as the maximizer of \(\tar(\obs_2)$. Then, find \(\meas_3\) by
%% keeping \(\meas_1,\meas_2\) fixed and taking \(\meas_3\) as the maximizer
%% of \(\tar(\obs_3)$. Continue this way until \(\obs_m = \obs\) is
%% found, where \(m\) is the number of available observations. %% It is
%% %% important to notice that this scheme does not require actually
%% %% observing data --- in \eqref{eq:objective} data is averaged out.

%% The analysis in this paper is conducted for the general simultaneous
%% optimization case. The sequential optimization case is dealt with in
%% section \ref{subsec:clusterization_sequential}. It is important to
%% note, however that all conclusions we arrive at for the simultaneous
%% case easily specialize to the sequential case by considering the
%% posterior as the next sequential step's prior.

\section{The Constrained Optimization Problem of D-Optimal Design}\label{section:D_and_grad}
We seek a formulation of the D-optimal design problem via Lagrange
multipliers. We first find the gradient of $\tar$, then we suggest
unit-norm constraints on $\obs$ and find their gradients. Results of
this section are summarized in Theorem \ref{thm:constrained}. First,
recall that:
\begin{definition}\label{def:var}
  Let $F$ a real valued function of $\obs$. The first variation of $F$
  at $\obs$ in the direction $V$ is:
  \begin{equation*}
    \delta F(\obs) V := \frac{\der}{\der \tau}\Big |_{\tau=0}  F( \obs + \tau V).
  \end{equation*}

@@

\begin{proposition}\label{prop:tar_grad}
  The gradient of the D-optimality objective $\tar$ is
  \begin{equation*}
    %% \delta \tar(\obs) V = \tr{V ( I - \modcov \obs^* \Sigma^{-1}\obs )
    %%   \fwd \postcov \fwd^* \obs^* \Sigma^{-1}}.
    \nabla \tar(\obs) = ( I - \modcov \obs^* \Sigma^{-1}\obs ) \fwd
    \postcov \fwd^* \obs^* \Sigma^{-1}
  \end{equation*}
\end{proposition}

@@

  \begin{align}\label{eq:der_sig}
    \begin{split}
      \frac{\der}{\der \tau} \Big |_{\tau=0} \Sigma( \obs + \tau V )
      &= \frac{\der}{\der \tau} \Big |_{\tau=0} 
      (\obs + \tau V ) \modcov (\obs + \tau V )^*  + \sigma^2I\\
      % 
      % 
      % 
      &= V \modcov \obs^* + \obs \modcov V^*.
    \end{split}
  \end{align}

  Then, using \eqref{eq:der_sig}: 
  \begin{align*}
    0 &= \frac{\der}{\der \tau} \Big |_{\tau=0} I \\
    % 
    % 
    % 
    &= \frac{\der}{\der \tau} \Big |_{\tau=0}
    \left (\Sigma(\obs+\tau V)^{-1} \Sigma(\obs+\tau V) \right ) \\
    % 
    % 
    % 
    &= \frac{\der \Sigma(\obs+\tau V)^{-1}}{\der \tau} \Big |_{\tau=0} \Sigma+
    \Sigma^{-1} \frac{\der \Sigma(\obs+\tau V)}{\der \tau} \Big |_{\tau=0}\\  
    %
    %
    %
    &= \frac{\der \Sigma(\obs+\tau V)^{-1}}{\der \tau} \Big |_{\tau=0} \Sigma+
    \Sigma^{-1} (V\modcov \obs^* + \obs \modcov V^*). 
    %\text{, by \eqref{eq:der_sig}. }
  \end{align*}

  Thus:
  \begin{align}\label{eq:der_sig_inv}
    \frac{\der \Sigma(\obs+\tau V)^{-1}}{\der \tau} \Big |_{\tau=0}  
      &= -\Sigma^{-1} (V \modcov \obs^* + \obs \modcov V^*) \Sigma^{-1}.
    \end{align}
  %

@@

  By Leibniz (product) rule and \eqref{eq:der_sig_inv}:
  % 
  \begin{align}\label{eq:T}
    \begin{split}
    \delta T(\obs) V 
    &= \frac{\der T(\obs + \tau V)}{\der \tau} \Big |_{\tau=0} \\
    %
    %
    %
    &= V^* \Sigma^{-1} \obs 
    - \obs^*\Sigma^{-1} V\modcov \obs^* \Sigma^{-1}\obs \\
    &\ \ \ - \obs^* \Sigma^{-1} \obs \modcov V^* \Sigma^{-1}\obs
    + \obs^* \Sigma^{-1} V.
    \end{split}
  \end{align}

@@

  \begin{align*}
    %\begin{split}
      \delta \tar(\obs) V 
      :&= \frac{\der}{\der\tau} \Big |_{\tau=0} \tar(\obs + \tau V) \text{ (Definition \ref{def:var})}\\
      % 
      % 
      % 
      &= \frac12 \frac{\der}{\der \tau} \Big |_{\tau=0} \log \det 
      (I + \prcov^{1/2} \fwd^* T(\obs+\tau V)\fwd \prcov^{1/2} ) \text{ (Theorem \ref{thm:d_optimality})} \\
      % 
      % 
      % 
      &= \frac12 \tr{( I + \prcov^{1/2} \fwd^* \obs^* \Sigma^{-1}
        \obs\fwd \prcov^{1/2} )^{-1}
        \frac{\der}{\der \tau} \Big |_{\tau=0}
        \prcov^{1/2} \fwd^* T(\obs+\tau V) \fwd \prcov^{1/2}}\ \text{ (Lemma \ref{lemma:lax})} \\
      % 
      % 
      % 
      &= \frac12 \ttr\Big \{ \postcov \fwd^* (V^* \Sigma^{-1} \obs 
      - \obs^*\Sigma^{-1} V\modcov \obs^* \Sigma^{-1}\obs \\
      &\ \ \ - \obs^* \Sigma^{-1} \obs \modcov V^* \Sigma^{-1}\obs 
      + \obs^* \Sigma^{-1} V ) \fwd \Big \}  \text{ (by \eqref{eq:T})} \\
      %
      %
      %
      &= \tr{\postcov \fwd^* ( \obs^* \Sigma^{-1} V -
      \obs^*\Sigma^{-1} V\modcov \obs^* \Sigma^{-1}\obs ) \fwd} \\
      %
      %
      % 
      &= \tr{\postcov \fwd^* \obs^* \Sigma^{-1} V 
      ( I - \modcov \obs^* \Sigma^{-1}\obs ) \fwd} \\
      % 
      %
      %
      &= \tr{V ( I - \modcov \obs^* \Sigma^{-1}\obs )
      \fwd \postcov \fwd^* \obs^* \Sigma^{-1}}.
    %\end{split}
  \end{align*} 
  Recalling Definition \ref{def:var} concludes the proof.
\end{proof}

@@

\begin{proposition}\label{prop:bigger_better}
  Let $\obs = (\meas_1,\dots,\meas_m)^t$, $j \in \{1,\dots,m\}$,
  $\sigma^2 > 0$ and $|\zeta| > 1$. Then $\tar(\obs)$ increases if we
  use $\zeta \meas_j$ in $\obs$ instead of $\meas_j$.
\end{proposition}

\begin{proof} 
  Fix $j=1,\dots,m$ and take $V:= e_j e_j^t \obs$. For $u
  \in \hilo$:
  \begin{equation*}
    Vu = e_je_j^t (\meas_1(u),\dots,\meas_m(u) )^t = e_j \meas_j(u)
    = (0,\dots,0,\meas_j(u),0,\dots,0)^t.
  \end{equation*}
  %% This way, $V$ has the same $j$th entry as $\obs$ while the rest
  %% are set to zero.
  We now calculate the variation of $\tar$ at $\obs$ in the direction
  of $V$. Denote $\tmp: = \fwd \postcov \fwd^*$. From Proposition
  \ref{prop:tar_grad}:
  \begin{align*}
     \delta \tar(\obs) V 
    &= \tr{V ( I - \modcov \obs^*\Sigma^{-1}\obs) \tmp \obs^* \Sigma^{-1}} \\
    % 
    %
    %
    &= \tr{e_je_j^t \obs ( I - \modcov \obs^*\Sigma^{-1}\obs) \tmp \obs^* \Sigma^{-1}} \\
    %
    % 
    %
    &= e_j^t \obs ( I - \modcov \obs^*\Sigma^{-1}\obs) \tmp \obs^* \Sigma^{-1}e_j \\
    %
    % 
    %
    &= e_j^t ( I - \obs \modcov \obs^*\Sigma^{-1})\obs \tmp \obs^* \Sigma^{-1}e_j \\  
    % 
    %
    %
    &=  e_j^t(\Sigma-\obs \modcov \obs^*) \Sigma^{-1}\obs \tmp \obs^* \Sigma^{-1}e_j \\
    %
    %
    %
    &=\sigma^2 e_j^t \Sigma^{-1}\obs \tmp \obs^* \Sigma^{-1}e_j
    \text{ by \eqref{eq:Sigma} }\\
    %
    % 
    %
    &=\sigma^2 e_j^t \Sigma^{-1}\obs \fwd \postcov \fwd^* \obs^* \Sigma^{-1}e_j.
  \end{align*} 
  Since $\postcov$ is positive definite, we conclude that $\delta
  \tar(\obs) V > 0$. This means that increasing the magnitude of the
  $j^{\text{th}}$ measurement functional increases $\tar(\obs)$.
\end{proof}

Proposition \ref{prop:bigger_better} implies that it is a good idea to
bound the norm of measurements. If, for example, we can take
measurements in $span\{\meas\}$ for some $\meas \neq 0$, then a
maximum for the D-optimality criterion does not exist, as it is not
bounded. In contrast, in any real-life problem where sensors are
concerned, the norm of measurements recorded by sensors is always one
(of course, point evaluations are not in any Hilbert space of
functions we wish to consider):
